Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap

### 2017-07-27 TeX/conference_paper.tex
finished paper revisions

\begin{abstract}
\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 1,000 nodes (i.e., one million potential ties) or greater. When the network is large, one option is maximum pseudolikelihood estimation (MPLE). The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. We find that bootstrapped MPLE can be run in 1/10th the time of MCMLE. We study the relative performance of MCMLE and MPLE with simulation studies, and illustrate the two different approaches by applying them to a network of bills introduced in the United State Senate.

@@

\section{Estimation}
\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.  Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible. The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}. Under maximum pseudolikelihood, the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.  
%
\begin{equation*}
\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
%\label{ILogit}
\end{equation*}
where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
\begin{equation*}
\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
%\label{ILogit}
\end{equation*}
where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}.

This corresponds with the \textit{logistic regression} approach, where the observations of the dependent variables are simply edge values of the observed adjacency matrix, and the observations of the covariate values are given as the scores of every single change statistic. Therefore, the estimation of $\theta$ can then be obtained as usual using straight forward maximum-likelihood estimation. The resulting likelihood function is of the following form:
\begin{equation}
\text{lik}(\theta)= \mathbb{P}_{\theta}(Y=G)= \prod_{i,j} \dfrac{ \exp \left(\theta^T \Delta(G))_{ij} \right)}{1+\exp \left(\theta^T \Delta(G))_{ij} \right)}
\label{NaiveLik}
\end{equation}
The problem with this simple estimation procedure is that the assumed hypothesis of the independence of the dyads turns out to be erroneous in many cases (see van Duijn et al. \cite{vanDuijnetal2009}). This is a systematic problem: The presence of network data is inextricably connected with the presence of \textit{relational data}, which by definition should not be assumed to be independent of each other. If this dependency structure is deliberately ignored and equation (\ref{NaiveLik}) is used to estimate $\theta$, it results in a \textit{maximum pseudo-likelihood estimation} (MPLE). This technique tends to underestimate the standard error. However, Desmarais and Cranmer \cite{Desmarais.2012} argue that the pseudo-likelihood provides a consistent approximation of the maximum likelihood, meaning that the MPLE converges to the MLE as the size of the network increases.\\[0.3cm]
There are several techniques to circumvent estimators, which underestimate the standard error of $\theta$. In the following, we will introduce a technique based on \textit{Markov Chain Monte Carlo (MCMC)} and maximum-likelihood methods.\\[0.3cm]
The more rigorous technique is to estimate the parameters directly with the log-likelihood function derived from (\ref{ERGM}), which has the following form:
%
\begin{equation}
\text{loglik}(\theta)=\theta^T \cdot \Gamma(G)-\log(c(\theta))
\label{loglik}
\end{equation}
%
where $G$ is the observed network. For the vector of network statistics, one can assume without loss of generality
%
\begin{equation}
\Gamma(G)=0 \label{assumption}
\end{equation}
%
This means that centering the vector of network statistics does not affect the probability function of the network variable $Y$. Therefore, in context of the likelihood function (\ref{loglik}) the vector of statistics can always be assumed to be centered on the observed network.\\
Due to assumption (\ref{assumption}), one gets from (\ref{loglik}) the simplified log-likelihood function
%
\begin{equation}
\text{loglik}(\theta)= -\log(c(\theta))
\label{vloglik}
\end{equation}
%
The problem resulting from estimating the parameters with (\ref{loglik}) is that the term
%
$$c(\theta):= \sum_{G^* \in \mathcal{G}(N_V)} \exp(\theta^T \cdot \Gamma(G^*))$$ 
%
which sums up the weighted network statistics of all possible networks of $N$ nodes, has to be evaluated. Even for networks with small numbers of nodes this presents an enormous computational obstacle, and the necessary calculations for larger networks cannot currently be completed in any reasonable timeframe.
As a result, for sufficiently large networks it is not possible to estimate the parameters directly with the likelihood function.
\\[0.3cm]
An expedient for this limitation is based on the following consideration: Fix any vector of parameters $\theta_0 \in \Theta$ from the underlying parameter range $\Theta$ and compute for $\theta \in \Theta$ the expected value. Then, one can show that
%
\mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]=\frac{c(\theta)}{c(\theta_0)}
%
This equation offers the following possibility: If one draws $L$ random networks $G_1, \dots ,G_L$ out of a distribution $\mathbb{P}_{\theta_0}$ appropriately, one gets with the \textit{law of big numbers} the following relation:
%
\begin{equation}\label{konver}
\frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right) &
~~\stackrel {{\tiny\mathbb{P}_{\theta_0}}}{\longrightarrow}~~\\ \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right]& = \frac{c(\theta)}{c(\theta_0)}
\end{equation}
%
For a big enough number, $L$, of random networks, the following approximation is reasonable:
%
\begin{equation}
\frac{c(\theta)}{c(\theta_0)} \approx \frac{1}{L}\cdot \sum_{i=1}^{L}  \exp\left((\theta - \theta_0)^T \cdot \Gamma(G_i)\right)
\label{approx}
\end{equation}
%
One can now use equation (\ref{approx}) to determine an approximation of the log-likelihood function (\ref{vloglik}):
%
\begin{eqnarray*}
&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\end{eqnarray*}
%
By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
%
\begin{equation}
s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\label{score}
\end{equation}
%
This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}. As a result, the approximate maximum-likelihood estimator for the parameters can be computed.\\[0.3cm]
As pleasant as this may sound, the immediate question arises: How can a sufficient number of suitable drawings $G_1,...,G_L$ be taken from the distribution $\mathbb{P}_{\theta_0}$? \\
For this purpose, the \textit{Markov Chain Monte Carlo (MCMC)} methods can be used. This approach does not deliberately ignore the dependency structures inside the network. Furthermore, the Markov Chain Monte Carlo Maximum Likelihood Estimator (MCMLE) approaches the MLE as the number of networks simulated to approximate the likelihood goes to infinity.\\[0.3cm]
No matter which kind of MCMC algorithm is used, the basic idea is the following:
\\One constructs a \textit{Markov chain} $(G_t)_{t \in \mathbb{N}}$ on the set of all possible networks $\mathcal{G}(N)$ of $N$ nodes, whose \textit{stationary distribution} is in conformity with the distribution $\mathbb{P}_{\theta_0}$. One can show that every single realization (or \textit{trajectory}) 
of this stochastic process accomplishes the convergence result (\ref{konver}) (for this version of the \textit{Law of big numbers for Markov chains} we reference Meyn and Tweedie \cite{Meyn.2009}). As a result, sub-sequences of $(G_t)_{t \in \mathbb{N}}$ which are sufficiently large enough can be used for approximation (\ref{approx}).\\[0.3cm]
Snijders \cite{Snijders.2002} provides a Metropolis Hastings algorithm on how to simulate networks:
Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
\begin{enumerate}
\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
\item Compute the value
$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
\begin{itemize}
\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
\item $Z=1$, define $G^{(k+1)}$ via
$$g_{pq}^{(k+1)}=\begin{cases}
1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
\end{cases}$$
\end{itemize}
\item Start at step 1 with $G^{(k+1)}$.
\end{enumerate}
\vspace{0.3cm}
\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$. Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.

\section{Efficiency of MPLE and MCMLE}
\noindent Even though the MCMLE is in general favored over the MPLE there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods. As mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\[0.3cm]
At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework. They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network. 
The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e. instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases. \\[0.3cm]%This means that the MPLE of a network on $1000$ nodes is closer to the true coefficient as the MPLE of a network on $100$ nodes. \\
%A major advantage of the MCMLE method is, as shown by van Duijn et al. \cite{vanDuijnetal2009}, that the MCMLE is substantially more efficient than the MPLE, meaning that the variance of the MCMLE is in general lower than the variance of the MPLE. Nevertheless, as shown by 
Next, we are going to compare the efficiency of the MCMLE compared to the MPLE. Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network increases. This means that for very large networks it becomes difficult to determine the number of simulated networks required in order for the MCMLE to outperform the MPLE. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMLE approach.\\[0.3cm]
\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al. \cite{Resnicketal1997}.\\[0.3cm]
For both networks we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$. Then, we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) were we fix the decay parameter $\lambda$ at $0.25$.
The number of edges in these unidrected networks are defined as

@@

In order to include nodal covariates into the ERGM, the vector of nodal attributes is expanded into an artificial matrix $C$, which has the same dimensions as $G$. The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times. The second row of matrix $C$, consists of the second actor's attribute, repeated $N$ times, and so on. Then, the statistics for a nodal covariate is defined as

@@

A precise description of the gwesp statistic will be given in the chapter VI.\\[0.3cm]
We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks. For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10.000$ simulated networks used in the likelihood approximation.
Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 

@@

\noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks the larger Faux Magnolia High network already requires a sample size of at least 1500 networks for the MCMLE to surpass the MPLE. These results propose especially for very large networks (e.g. social media data) the question, of how large the sample size has to be set in order to justify the approximately exact, but computationally expensive MCMLE method. The identification of the required sample size is not only becoming more difficult as the size of the network increases, but an increase of the sample size will also extend the mere computation time.

\section{Bootstrapped MPLE}
\noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size is not chosen large enough. The main reason why the MCMLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the standard deviation (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. For exponential family distributions it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE. \\[0.3cm]
\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network. For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$. For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals. Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval. 
\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{Coverage_Mesa_Magnolia_height05}
\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression}
\label{coverage}
\end{figure}

>>>>

\begin{abstract}
\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 1,000 nodes (i.e., one million potential ties) or greater. When the network is large, one option is maximum pseudolikelihood estimation (MPLE). The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. We find that bootstrapped MPLE can be run in 1/5th the time of MCMLE. We study the relative performance of MCMLE and MPLE with simulation studies, and illustrate the two different approaches by applying them to a network of bills introduced in the United State Senate.

@@

\section{Estimation}
\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.  Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible. The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}. Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.  
%
\begin{equation*}
\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
%\label{ILogit}
\end{equation*}
where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
\begin{equation*}
\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
%\label{ILogit}
\end{equation*}
where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}. For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.

Despite the computational efficiency underlying the implementation of MPLE,  existing methods for assessing uncertainty with respect to the MPLE perform poorly (see van Duijn et al. \cite{vanDuijnetal2009}). Estimating the uncertainty in parameter estimates (e.g., standard errors, confidence intervals), is a critical step in using the results from a statistical model. Estimates of uncertainty are used to test hypotheses about parameters, estimate variance (i.e., risk) in model predictions, and estimate effect sizes. Another limitation of MPLE regards the assessment of model fit. With ERGM, model fit is conventionally evaluated by comparing the structure of the observed network to networks simulated from the estimated model \cite{Hunter.2008}. This comparison can include diagnosing the model for a particular form of poor fit that arises with ERGM---model degeneracy \cite{schweinberger2011instability}. Degeneracy is a condition in which the model places nearly all of the probability mass on the completely empty or completely full network. Since MPLE does not require simulating networks, researchers can derive and report results without checking the fit of the model or checking for degeneracy. Of course, the researcher can choose to simulate networks and check model fit with MPLE , but unlike the simulation-based methods of estimation, it is possible to run MPLE without simulating networks.

The contemporary conventional approach to estimating $\theta$, introduced by Snijders (2002) \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\[0.3cm]
This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}). The log-likelihood function is not evaluated directly, rather, the log ratio of the likelihood under a proposed value of the parameters $\theta$, and an initial value of the parameters $\theta_0$, is approximated using $L$ networks simulated from the ERGM with parameter values $\theta_0$. The approximation, detailed in Snijders (2002) \cite{snijders2002markov} is given by 
%
\begin{eqnarray*}
&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
\end{eqnarray*}
By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
%
\begin{equation}
s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\label{score}
\end{equation}
%
This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.

MCMC methods are used to simulate the $L$ networks. As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. Snijders (2002) \cite{Snijders.2002} provides a Metropolis Hastings algorithm to simulate networks:
Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
\begin{enumerate}
\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
\item Compute the value
$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
\begin{itemize}
\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
\item $Z=1$, define $G^{(k+1)}$ via
$$g_{pq}^{(k+1)}=\begin{cases}
1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
\end{cases}$$
\end{itemize}
\item Start at step 1 with $G^{(k+1)}$.
\end{enumerate}
\vspace{0.3cm}
\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}. Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.

\section{Efficiency of MPLE and MCMLE}
\noindent  As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time. At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework. They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network.  The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e. instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.

To illustrate the relative efficiency of MPLE and MCMLE Next, we run a simulation study. Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood in MCMLE is not large enough. It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network (i.e., the number of nodes in the network) increases. This means that, for very large networks, it becomes difficult to determine the number of simulated networks required for the MCMLE to outperform the MPLE. In other words, the larger the network of the more computationally intensive it becomes to use MCMLE in a way that out-performs MPLE.\\[0.3cm]
\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's \cite{HunterDavidR..2008} Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al. \cite{Resnicketal1997}.\\[0.3cm]
For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$. Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.
The number of edges is defined as

@@

In order to include nodal covariates into the ERGM, the vector of nodal attribute is expanded into a matrix $C$, which has the same dimensions as $G$. The first row of matrix $C$ consists of the first actor's attribute, repeated $N$ times. The second row of matrix $C$, consists of the second actor's attribute, repeated $N$ times, and so on. Then, the statistic for a nodal covariate is defined as

@@

The GWESP statistic is given as
\Gamma_{esp}: \mathcal{G}(N) \to \mathbb{R}~~,~~G \mapsto \sum_{i<j}^{N}\mathds{1}_k\biggl(\sum_{k=1}^N g_{ij}g_{im}g_{jm}\biggr).
$\Gamma_{esp}(G,k)$ counts the number of nodal pairs $(i,j)$, which share exactly $k$ neighbors. This statistic is used to model the tendency towards triangles and clustering in a network.\\
We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks. For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10,000$ simulated networks used in the likelihood approximation.
Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 

@@

\noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks, the larger Faux Magnolia High network requires a sample size of at least 1,500 networks for the MCMLE to surpass the MPLE.  For especially large networks (e.g., social media data) the sample size has to be set in order to justify the approximately exact, but computationally expensive, MCMLE method, may be prohibitive.

\section{Bootstrapped MPLE}

\noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size used in MCMLE is not large enough. The main reason why the MCMLE is still widely preferred is that, in contrast to the MPLE, it does not underestimate the standard errors (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (G)$ is the sufficient statistic. For exponential family distributions, it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the nominal 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as bootstrap resampling \cite{efron1982jackknife}. Bootstrap resampling refers to constructing a sampling distribution for the parameter estimate by resampling the data with replacement, and re-estimating the model on the resampled data. Under non-parametric bootstrap resampling, the data are resampled directly from the dataset. Under the parametric boostrap, the data is resampled from the estimated model. The idea of using boostrap resampling with MPLE for ERGM was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that the MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that bootstrap resampling consistently estimates the confidence intervals of the MPLE. \\[0.3cm]
\noindent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap. The non-parametric bootstrap cannot be applied when there is just a single network under study.

For the case in which there is just a single network to be studied, which is indeed the most common case in the literature, we propose the use of a parametric bootstrap. Under the parametric bootstrap, the sampling distribution of the MPLE is derived by re-estimating the MPLE on a sample of networks simulated from the MPLE estimated on the observed network. We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same parametrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network.

For the simulation study, we determine the MPLE for the model and treat these estimates as the networks' 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $G$. For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain  95\% bootstrap confidence intervals. Similar as for the MCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval. 
\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{Coverage_Mesa_Magnolia_height05}
\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMLE and logistic regression}
\label{coverage}
\end{figure}
